For submission to the Journal of Integrated Pest Management

### 2021-08-23 ofv_fife_2021.Rmd
Added accession numbers, changed mentions of 'strain' to 'isolate' and added a link to the 'Virus Detection' section in the table, updated the table title and footnote

We describe the first detection of orchid fleck virus (OFV), belonging to the orchid-infecting subgroup (OFV-Orc), infecting three unreported hosts: *Liriope muscari*, cv. 'Gigantea' (Decaisne) Bailey, *Ophiopogon intermedius* Don and *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea) in Leon and Alachua Counties, FL. Strains of OFV-Orc infect over 50 plant species belonging to the plant families Orchidaceae, Asparagaceae (Nolinoidaea), and infects *Citrus* (Rutaceae) as citrus leprosis disease. The only known vectors of OFV-Orc are the flat mites *Brevipalpus californicus* (Banks) *sensu lato* (Trombidiformes: Tenuipalpidae). Florida has various plants in the landscape which *Brevipalpus* spp. feed on, which are susceptible to infection by OFV-Orc.

@@

Two orchid-infecting strains of OFV were detected via combinations of conventional RT-PCR, RT-qPCR, Sanger sequencing and High Throughput Sequencing. Amplicons shared 98% nucleotide identity with OFV-Orc1 and OFV-Orc2 RNA2 genome sequences available in NCBI GenBank. Coinfections were seen in each county, but single strains of OFV-Orc were seen in *L. muscari* (Alachua, OFV-Orc2) and *A. elatior* (Leon, OFV-Orc1).

@@

OFV-infected plants exhibit various symptoms depending on the infected plant species as well as the strain of the OFV associated with the infection [@Kubo2009b], but symptoms typically appear as chlorotic flecks, which ultimately coalesce into ringspot patterns.

@@

The FDACS determined that the pathogen was OFV using previously published primers and methods to conduct RT-PCR and Sanger sequencing [@Kubo2009a; @Kubo2009b; @RamosGonzalez2015]. Orchid subgroup 1, OFV-Orc, was identified following the methods described in @Kondo2017. Sequencing demonstrated a shared 98\% nucleotide identity with the orchid strain subgroup, OFV-Orc (isolates So and Br with GenBank Accession numbers: AB244418 and MK522807, respectively) [@Kondo2006; @Kondo2017].

These samples from FDACS were subsequently retested by the USDA-APHIS-PPQ S&T Beltsville laboratory, in conjunction with tests of fresh samples from both Alachua and Leon counties. The USDA used RT-PCR, RT-qPCR, and High Throughput Sequencing (HTS) to reconfirm the presence of OFV. RT-PCR and qPCR with Generic R2-Dicho-GF and R2-Dicho-GR primers amplifed ~800 nt of L-gene (RNA2) amplicon [@Roy2020], and OFV-Orc1 and OFV-Orc2 were detected in both *O. intermedius* and *A. elatior* from Leon County. OFV-Orc1 and OFV-Orc2 strains share >90% nt sequence identity and differ >25% in the L gene nucleotide sequence, indicating two separate species of *Dichoravirus* [@Dietzgen2014]<!-- GenBank Accession No# and No #, respectively-->.

HTS reaffirmed the presence of OFV-Orc1 and OFV-Orc2 strains in Leon and Alachua counties (`r citet("tab_1")`). HTS results from Leon County revealed that *L. muscari* were coinfected with both strains (OFV-Orc1 and OFV-Orc2), while *A. elatior* were solely infected with OFV-Orc1. HTS of *L. muscari* from Alachua County revealed infections with the OFV-Orc2 strain.

After the initial identification by FDACS of OFV-Orc strains, mite samples were collected from symptomatic Asparagaceae in Leon County. Most mites collected were Tenuipalpid mites (flat mites or false spider mites), a pest of ornamental plants, some of which are known to act as vectors for plant viruses [@Childers2003; @Childers2011].

@@

OFV consists of two orchid strains (OFV-Orc1 and OFV-Orc2) and two citrus strains (OFV-Cit1 and OFV-Cit2) [@Roy2020; @Beltran-Beltran2020]. The OFV strains detected in Florida are identical in gene order, content, and genome sequence to the orchid strains of OFV infecting citrus in Hawaii, Mexico, Colombia, and South Africa [@Roy2020; @Beltran-Beltran2020]. Both OFV-Orc1 and OFV-Orc2 infect citrus [@Roy2020], but none of the citrus strains have been reported from any orchid species. The *Brevipalpus* mites collected from liriopogons and *A. elatior* in Leon County were abundant on OFV-infected plants very near to citrus trees, some plants even surrounding the trunk. *B. californicus* s. l. has been reported as a pest of citrus [@Childers2003] and are often collected from citrus fruits [@Vacante2016; @Vacante2010; @Baker1987; @Baker1949]. The proximity of these mite vectors to citrus raises the question: why these trees are not currently infected with OFV-Orc? It is important to note the uncertainty surrounding the vector for OFV-Orc. There are three mite species which have been recovered from OFV-Orc infected plants: *B. obovatus*, and *B. confusus* and *B. californicus* s.l., but only *B. californicus* has been described as a vector of OFV. Even so, the *B. californicus* which we find on liriopogons and *A. elatior* may not be the same cryptic species as those found on citrus. Transmission of OFV from populations of *B. californicus* liriopogon/*A. elatior* to citrus may be limited by host preferences, vectorial capacity, viral propagation/circulation in the vector, viral acquisition times, or feeding times required for transmission to citrus. Even so, these types of questions require future study to determine the potential of nolinoidaea to citrus transmission.

@@

In conclusion, detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivated *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. Citrus leprosis was present in Florida during the 1860’s and almost eradicated by the mid-1960s [@Childers2003; @Knorr1968a; @Knorr1968b]. An examination of herbarium specimens of Florida citrus found that this historical virus, Citrus leprosis dichorhavirus-N0, is distantly related to the modern strains of OFV [@Hartung2015; @Kitajima2011a; @Roy2020]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) associated with leprosis-like symptoms highlights the potential threat of different strains of OFV on citrus, which will be a definite concern to the US multi-billion-dollar citrus industry already impacted by the Huanglongbing disease. *B. californicus*, *B. yothersi*, and *B. obovatus* are all present in Florida [@Childers2003; @Akyazi2017], and are difficult to identify by non-experts, or without advanced methodologies. DNA barcoding [@Armstrong2005] or a similarly simple and accurate method for identification of these mite complexes is vital to identify mite populations which need to be monitored or controlled. By doing so, we can determine the risk OFV-Orc represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.

### Acknowledgements
We would like to give special thanks for the Tallahassee Museum and their patience, cooperation, and support with collecting plant samples. We are grateful for the USDA-APHIS PPQ Beltsville laboratory for their support in the identification and confirmation OFV strains, as well as *Brevipalpus* mite identification. We also want to thank Drs. Sam Bolton, FDACS and Aline Tassi, Univ. of Sao Paulo, Brazil for checking the mites we have sent for species validation. Furthermore, we are grateful for Dr. Marc S. Frank's identification of the Liriopogons collected. We are especially indebted to the late Dr. Gary Bauchan for his contributions to this study and the field of acarology, he will be greatly missed. This research was partly funded by the USDA National Institute of Food and Agriculture, Hatch project FLA-NFC-005607. Mention of trade names or commercial products in this publication is solely for the purpose of providing specific information and does not imply recommendation or endorsement by the USDA; USDA is an equal opportunity provider and employer.

@@

### Table 1: List of Asparagaceae (Nolinoidaea) species with verified cases of orchid fleck virus, collected from the landscape of northern Florida

|Scientific Name|Common Names|County|Strains|
|:-----------|:--------------|:-----|:------|
|*Liriope muscari* cv. 'Gigantea'\* (Decaisne) Bailey | Lilyturf, Orchardgrass, Monkeygrass|Alachua & Leon|OFV-Orc1 & OFV-Orc2|
|*Ophiopogon intermedius*\*\* Don| Aztec Grass, 'Argenteomarginatus'|Leon|OFV-Orc1 & OFV-Orc2|
|*Aspidistra elatior* Blume| Cast Iron Plant, Bar-room Plant|Leon|OFV-Orc1 & OFV-Orc2|
`r tab_nums(name = "tab_1", caption = "* *Liriope muscari* cv. 'Gigantea' has been traditionally classified as *L. gigantea* Hume by @Broussard2007 and @Fantz2015, although this distinction has been challenged by @Wang2014 and @Masiero2020. * * *O. intermedius* is sometimes misclassified as *Liriope muscari* ‘Variegated Evergreen Giant’ [@Fantz2008b, @Fantz2009] or ‘Grandiflora White’ [@Fantz2009]. Both OFV-Orc1 and OFV-Orc2 were detected in all host species.")`

>>>>

We describe the first detection of orchid fleck virus (OFV), belonging to the orchid-infecting subgroup (OFV-Orc), infecting three unreported hosts: *Liriope muscari*, cv. 'Gigantea' (Decaisne) Bailey, *Ophiopogon intermedius* Don and *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea) in Leon and Alachua Counties, FL. OFV-Orc infects over 50 plant species belonging to the plant families Orchidaceae, Asparagaceae (Nolinoidaea), and infects *Citrus* (Rutaceae) as citrus leprosis disease. The only known vectors of OFV-Orc are the flat mites *Brevipalpus californicus* (Banks) *sensu lato* (Trombidiformes: Tenuipalpidae). Florida has various plants in the landscape which *Brevipalpus* spp. feed on, which are susceptible to infection by OFV-Orc.

@@

Two orchid-infecting isolates of OFV were detected via combinations of conventional RT-PCR, RT-qPCR, Sanger sequencing and High Throughput Sequencing. Amplicons shared 98% nucleotide identity with OFV-Orc1 and OFV-Orc2 RNA2 genome sequences available in NCBI GenBank. Coinfections were seen in each county, but single isolates of OFV-Orc were seen in *L. muscari* (Alachua, OFV-Orc2) and *A. elatior* (Leon, OFV-Orc1).

@@

OFV-infected plants exhibit various symptoms depending on the infected plant species as well as the isolate of the OFV associated with the infection [@Kubo2009b], but symptoms typically appear as chlorotic flecks, which ultimately coalesce into ringspot patterns.

@@

The FDACS determined that the pathogen was OFV using previously published primers and methods to conduct RT-PCR and Sanger sequencing [@Kubo2009a; @Kubo2009b; @RamosGonzalez2015]. Orchid subgroup 1, OFV-Orc, was identified following the methods described in @Kondo2017. Sequencing demonstrated a shared 98\% nucleotide identity with the orchid isolate subgroup, OFV-Orc (isolates So and Br with GenBank Accession numbers: AB244418 and MK522807, respectively) [@Kondo2006; @Kondo2017].

These samples from FDACS were subsequently retested by the USDA-APHIS-PPQ S&T Beltsville laboratory, in conjunction with tests of fresh samples from both Alachua and Leon counties. The USDA used RT-PCR, RT-qPCR, and High Throughput Sequencing (HTS) to reconfirm the presence of OFV. RT-PCR and qPCR with Generic R2-Dicho-GF and R2-Dicho-GR primers amplifed ~800 nt of L-gene (RNA2) amplicon [@Roy2020], and OFV-Orc1 and OFV-Orc2 were detected in both *O. intermedius* and *A. elatior* from Leon County (GenBank Accession Numbers: MZ852004, MZ852005 MZ852006, and MZ852007). OFV-Orc1 and OFV-Orc2 share >90% nt sequence identity and differ >25% in the L gene nucleotide sequence, indicating two separate species of *Dichoravirus* [@Dietzgen2014].

HTS reaffirmed the presence of OFV-Orc1 and OFV-Orc2 in Leon and Alachua counties (`r citet("tab_1")`). HTS results from Leon County revealed that *L. muscari* were coinfected with both OFV-Orc1 and OFV-Orc2, while *A. elatior* were solely infected with OFV-Orc1. HTS of *L. muscari* from Alachua County revealed infections with OFV-Orc2 (GenBank Accession MZ852006).

After the initial identification by FDACS of OFV-Orc, mite samples were collected from symptomatic Asparagaceae in Leon County. Most mites collected were Tenuipalpid mites (flat mites or false spider mites), a pest of ornamental plants, some of which are known to act as vectors for plant viruses [@Childers2003; @Childers2011].

@@

OFV consists of two orchid isolates (OFV-Orc1 and OFV-Orc2) and two citrus isolates (OFV-Cit1 and OFV-Cit2) [@Roy2020; @Beltran-Beltran2020]. The OFV isolates detected in Florida are identical in gene order, content, and genome sequence to the orchid isolates of OFV infecting citrus in Hawaii, Mexico, Colombia, and South Africa [@Roy2020; @Beltran-Beltran2020]. Both OFV-Orc1 and OFV-Orc2 infect citrus [@Roy2020], but none of the citrus  isolates have been reported from any orchid species. The *Brevipalpus* mites collected from liriopogons and *A. elatior* in Leon County were abundant on OFV-infected plants very near to citrus trees, some plants even surrounding the trunk. *B. californicus* s. l. has been reported as a pest of citrus [@Childers2003] and are often collected from citrus fruits [@Vacante2016; @Vacante2010; @Baker1987; @Baker1949]. The proximity of these mite vectors to citrus raises the question: why these trees are not currently infected with OFV-Orc? It is important to note the uncertainty surrounding the vector for OFV-Orc. There are three mite species which have been recovered from OFV-Orc infected plants: *B. obovatus*, and *B. confusus* and *B. californicus* s.l., but only *B. californicus* has been described as a vector of OFV. Even so, the *B. californicus* which we find on liriopogons and *A. elatior* may not be the same cryptic species as those found on citrus. Transmission of OFV from populations of *B. californicus* liriopogon/*A. elatior* to citrus may be limited by host preferences, vectorial capacity, viral propagation/circulation in the vector, viral acquisition times, or feeding times required for transmission to citrus. Even so, these types of questions require future study to determine the potential of nolinoidaea to citrus transmission.

@@

In conclusion, detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivated *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. Citrus leprosis was present in Florida during the 1860’s and almost eradicated by the mid-1960s [@Childers2003; @Knorr1968a; @Knorr1968b]. An examination of herbarium specimens of Florida citrus found that this historical virus, Citrus leprosis dichorhavirus-N0, is distantly related to the modern isolates of OFV [@Hartung2015; @Kitajima2011a; @Roy2020]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) associated with leprosis-like symptoms highlights the potential threat of different isolates of OFV on citrus, which will be a definite concern to the US multi-billion-dollar citrus industry already impacted by the Huanglongbing disease. *B. californicus*, *B. yothersi*, and *B. obovatus* are all present in Florida [@Childers2003; @Akyazi2017], and are difficult to identify by non-experts, or without advanced methodologies. DNA barcoding [@Armstrong2005] or a similarly simple and accurate method for identification of these mite complexes is vital to identify mite populations which need to be monitored or controlled. By doing so, we can determine the risk OFV-Orc represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.

### Acknowledgements
We would like to give special thanks for the Tallahassee Museum and their patience, cooperation, and support with collecting plant samples. We are grateful for the USDA-APHIS PPQ Beltsville laboratory for their support in the identification and confirmation OFV isolates, as well as *Brevipalpus* mite identification. We also want to thank Drs. Sam Bolton, FDACS and Aline Tassi, Univ. of Sao Paulo, Brazil for checking the mites we have sent for species validation. Furthermore, we are grateful for Dr. Marc S. Frank's identification of the Liriopogons collected. We are especially indebted to the late Dr. Gary Bauchan for his contributions to this study and the field of acarology, he will be greatly missed. This research was partly funded by the USDA National Institute of Food and Agriculture, Hatch project FLA-NFC-005607. Mention of trade names or commercial products in this publication is solely for the purpose of providing specific information and does not imply recommendation or endorsement by the USDA; USDA is an equal opportunity provider and employer.

@@

### Table 1: List of Asparagaceae (Nolinoidaea) species infected with orchid fleck virus, collected from the landscape of northern Florida

|Scientific Name|Common Names|County|Isolates Detected|
|:-----------|:--------------|:-----|:------|
|*Liriope muscari* cv. 'Gigantea'\* (Decaisne) Bailey | Lilyturf, Orchardgrass, Monkeygrass|Alachua & Leon|OFV-Orc1 & OFV-Orc2|
|*Ophiopogon intermedius*\*\* Don| Aztec Grass, 'Argenteomarginatus'|Leon|OFV-Orc1 & OFV-Orc2|
|*Aspidistra elatior* Blume| Cast Iron Plant, Bar-room Plant|Leon|OFV-Orc1 & OFV-Orc2|
`r tab_nums(name = "tab_1", caption = "* *Liriope muscari* cv. 'Gigantea' has been traditionally classified as *L. gigantea* Hume by @Broussard2007 and @Fantz2015, although this distinction has been challenged by @Wang2014 and @Masiero2020. * * *O. intermedius* is sometimes misclassified as *Liriope muscari* ‘Variegated Evergreen Giant’ [@Fantz2008b, @Fantz2009] or ‘Grandiflora White’ [@Fantz2009]. Both OFV-Orc1 and OFV-Orc2 were detected in each species tested, many plants were coinfected with both isolates, see '[Virus Detection]'.")`
